---
title: "Essay_4"
format: pdf
editor: visual
---

# Wine Quality Prediction with KNN

### Authors: Preston O'Connor

### Date: 4/14/2025

## Introduction

```{r}
#libraries given in Step 1 of the slides
# install.packages("class") # For KNN
# install.packages("ggplot2") # For visualization
# library(class)
# library(ggplot2)

#loading the data set
```

## Data Description

### Check for Missing values

```{r}

```

### Box plots and Outliers

```{r}



```

### Normalize Data

```{r}
# normalize <- function(x) { (x - min(x)) / (max(x) - min(x)) }
# 
# iris_norm <- as.data.frame(lapply(iris_clean[, 1:4], normalize))
```

## Analysis

### Get Optimized K

```{r}
#found in step 4 of the knn in class Lecture slides for KNN
```

### Train and Test Split

```{r}
#found in step 2 of the knn in class Lecture slides for KNN
```

### KNN Implementation

```{r}
#found in step 3 of the knn in class Lecture slides for KNN
```

## Model Evaluation

### The Accuracy of the Model

```{r}
# accuracy <- sum(knn_final == test_labels) / length(test_labels)
# cat("Accuracy:", accuracy, "\n")
```

### The Confusion Matrix

```{r}
# conf_matrix <- confusionMatrix(knn_final, test_labels)
# print(conf_matrix)

```

### Evaluate the ROC curve and ACU

```{r}
# library(pROC)
# roc_curve <- roc(test_labels, as.numeric(knn_final))
# auc_value <- auc(roc_curve)
# cat("AUC:", auc_value, "\n")
```

## Conclusion & Summary

## Sources

-   Source 1

-   Source 2

-   Kaggle data set
